---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NMAR

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/ncn-foreigners/NMAR/graph/badge.svg)](https://app.codecov.io/gh/ncn-foreigners/NMAR)
[![R-CMD-check](https://github.com/ncn-foreigners/NMAR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ncn-foreigners/NMAR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of **NMAR** is to provide a set of functions to **estimate the population mean** of data subject to Not Missing At Random (NMAR) mechanisms using advanced statistical methods 

## Overview
The **NNMAR** library provides functions to estimate the mean of **Not Missing At Random (NMAR)** data. The estimation functions are built upon the following engine functions, each implementing a distinct method:

* `exptilt_engine`: Exponential Tilting Estimator (Riddles et al, 2016, [DOI:10.xxxx/xxxx](https://doi.org/10.xxxx/xxxx))
* `exptilt_nonparam_engine`: Nonparametric Exponential Tilting Estimator (Riddles et al, 2016, Appendix 2, [DOI:10.xxxx/xxxx](https://doi.org/10.xxxx/xxxx))
* `el_engine`: Empirical Likelihood Estimator (Kim and Yu, 2011, [DOI:10.xxxx/xxxx](https://doi.org/10.xxxx/xxxx))

The main user-facing function is `nmar()`, which acts as a unified wrapper around these engine functions.


## Installation
You can install the development version of NMAR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ncn-foreigners/NMAR")
```

or
```
remotes::install_github("ncn-foreigners/NMAR@main", force = T, build = T, build_manual = T, build_vignettes = T)
```

### Project branches:
- `main` - The stable production branch. This is the version you should install for general use
- `package-pre_prod` - The pre-production development branch. This is a stable, installable version reflecting the latest features. (1:1 clone of `package-dev`)
- `package-dev` - The **internal development branch**. This branch uses a non-R package structure for internal development purposes. **DO NOT** download or install this branch. For contribution guidelines, please see the CONTRIBUTING.md file or the [project webpage (TODO: add link to webpage)]

## Usage

### General naming
* `outcome_var` (f.e Y): The outcome variable (with missing values).
* `covariates_for_outcome`(f.e x1,x2): Predictors of `outcome_var` value (used in the response model).
* `covariates_for_missingness`(f.e x3): Predictor of `outcome_var` missingness (used in the missingness model).

`formula` = outcome_var ~ covariates_for_outcome

### Example

If, in a salary survey, richer people were less likely to answer, `Y` is salary, `x1` and `x2` are experience and education, and `x3` might be gender (if we assume gender affects the likelihood of responding but not the salary value itself).


```{r example}
generate_test_data <- function(n_rows = 400, n_cols = 2, case = 2, x_var = 0.5, eps_var = 0.5, a = 0.8, b = -0.2) {
  X <- as.data.frame(replicate(n_cols, rnorm(n_rows, 0, x_var)^1))
  colnames(X) <- paste0("x", 1:n_cols)
  coefs <- runif(n_cols, 0.95, 1.05)
  eps <- rnorm(n_rows, 0, eps_var)
  if (case == 1) {
  X$Y <- as.vector(-1 + as.matrix(X) %*% coefs + eps)
  }
  X <- X[order(X$Y), ]
  Y_original = X$Y
  pi_obs <- 1 / (1 + exp(-(a + b * X$Y)))
  mask <- runif(nrow(X)) > pi_obs
  mask[1] <- FALSE
  X$Y[mask] <- NA
  return(list(X = X, Y_original = Y_original))
}

library(NMAR, quietly = T)
res_test_data <- generate_test_data(n_rows = 500, n_cols = 3, case = 1)
data <- res_test_data$X

exptilt_config <- exptilt_engine(
  y_dens = 'normal',
  family = 'probit', # or logit
  variance_method='delta', #or delta
)

formula = Y ~ x1 + x2
res <- nmar(formula = formula, data = data, engine = exptilt_config, response_predictors = c('x3'))

print(res)
```


## Getting help
If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/ncn-foreigners/NMAR/issues)

## Authors and acknowledgments
Research grant: [OPUS 20 #2020/39/B/HS4/00941](#)

- [Maciej Beręsewicz](#)
- [Igor Kołodziej](#)
- [Mateusz Iwaniuk](#)

