---
title: "Exponential Tilting"
description: |
  Minsun Kim Riddles, Jae Kwang Kim, Jongho Im <br>
  A Propensity-score-adjustment Method for Nonignorable Nonresponse <br>
  https://doi.org/10.1093/jssam/smv047
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_exptilt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(NMAR)
```

# Overview
This vignette demonstrates the Parametric Exponential Tilting estimator for handling Not Missing at Random (NMAR). This method implements the maximum likelihood approach described in Riddles, Kim, and Im (2016)



# Input data
This method is applicable when:
- Data is individual-level with both the outcome variable and auxiliary covariates.
- The outcome variable follows a known distribution (e.g., normal).

## Required Structure
- **Outcome Variable**: A continuous variable subject to nonresponse (e.g., `y`).
- **Covariates for outcome model**: Variables trying to model Outcome Variable (e.g., `x`).
- **Instrumental Variable**: (optional) A variable affecting response probability but not directly related to the outcome.

```{r}
set.seed(1108)
head(riddles_case1)
```


```{r}
y_original <- riddles_case1$y_true # to compare with true values
```

# Engine Configuration
The estimation is configured using `exptilt_engine`. Critical parameters include the assumed outcome density for respondents and the variance estimation method

The model is specified via a formula: Outcome ~ Response_Covariates | Instrument.

- **LHS**: The outcome variable (Y) containing missing values.
- **RHS** (Left of |): Covariates included in the response propensity model .
- **RHS** (Right of |): The instrumental variable. The method assumes this variable predicts the outcome but is conditionally independent of the response status .

```{r}
# Exptilt engine configuration
exptilt_config <- exptilt_engine(
  y_dens = 'normal', # Assume f(y|x, delta=1) is Normal
  family = 'logit', # Logit link for response probability
  variance_method = 'bootstrap',
  bootstrap_reps = 5,
  control = list(maxit = 10), # Solver control parameters
  stopping_threshold = 0.01, # Convergence threshold
  standardize = FALSE
)

formula = y ~x
res <- nmar(formula = formula, data = riddles_case1, engine = exptilt_config, trace_level = 0)
```


# Results

The function returns an object containing the estimated coefficients for the response model and the adjusted mean estimate for the outcome.

The adjusted mean is calculated using propensity score weighting, where the weights are derived from the fitted nonignorable response model
```{r}
print(res)
```

```{r}
coef(res)
```

Comparing the results to the true population mean (known from the synthetic generation) demonstrates the bias correction.
```{r}
y_true_mean <- mean(riddles_case1$y_true)
est_mean <- as.numeric(res$y_hat)
se <- res$se

cat('Estimated (NMAR) Mean:', sprintf('%.4f', est_mean),
    ' (SE:', sprintf('%.4f', se), ')\n')
cat('True Y Mean:          ', sprintf('%.4f', y_true_mean), '\n')
cat('Naive (MCAR) Mean:     ', sprintf('%.4f', mean(riddles_case1$y, na.rm = TRUE)), '\n')
```
# Survey designs

The engine supports survey.design objects. When provided, the algorithm incorporates sampling weights into the likelihood estimation and the final propensity score weighting.

The weights affect both the estimation of the respondent density f(y∣x) and the solution to the mean score equation, ensuring population-level inference
```{r}
# if survey is installed
if (requireNamespace("survey", quietly = TRUE)) {

  library('survey')
  surv_test_weights <- runif(length(riddles_case1$y), 0.9, 1.1)
  surv_test_data <- svydesign(ids = ~1, data = riddles_case1, weights = ~surv_test_weights)

  exptilt_config <- exptilt_engine(
     standardize = FALSE,
     on_failure = "error",
     bootstrap_reps = 5,
     supress_warnings = FALSE,
     auxiliary_means = NULL,
     control = list(maxit = 15, method = "Newton"),
     family = "logit",
     y_dens = "normal",
     variance_method = "bootstrap",
     stopping_threshold = 1
)


  formula = y ~ x
  res <- nmar(formula = formula, data = surv_test_data, engine = exptilt_config, trace_level = 3)
  print(res)

}
```
